Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Edit figures in Results section manuscript.
Manuscript versie voor feedback Gerko

\title{Missing the Point\(\colon\) Non-Convergence in Iterative Imputation
Procedures}

@@

Missing data problems are ubiquitous and pervasive in the social and
behavioral sciences. If a dataset contains just one missing value for a
variable of interest, statistical inferences are undefined and will not
produce any result. To circumvent this, incomplete observations are
ignored by default in many statistical packages (i.e., list-wise
deletion is employed). Unfortunately, this \emph{ad hoc} solution may
yield wildly invalid results \citep{buur18}. An alternative is to
`impute' (i.e., fill in) every missing value in an incomplete dataset.
With imputation techniques, one or several completed datasets are
created, on which statistical inferences can be performed. The case with
several completed datasets is known as `multiple imputation'
\citep[MI;][]{rubin76}, and requires an additional step to pool the
results across imputations \textbf{see figure 1??}.

\textbf{Move this after notation??} Figure \ref{fig:diagram} provides an
overview of the steps involved with MI---from incomplete data, to \(M\)
completed datasets, to \(M\) estimated quantities of interest
(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
\(M\) completed datasets. On each completed dataset, the analysis of
scientific interest (the `substantive model', or `complete data model')
is performed. The quantity of scientific interest (e.g., a regression
coefficient) is denoted with \(Q\). Since \(Q\) is estimated on each
completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained. These
\(M\) values are combined into a single pooled estimate \(\bar{Q}\).
\textbf{ALSO: make clear that iteration happens within the first step,
between incomplete and completed data.}

@@

By applying Rubin's rules, the pooled result will reflect the
uncertainty in the data due to missingness. Under many circumstances,
this yields an unbiased and confidence valid estimate of the inference
on the true--but missing--data \citep{buur18}. A popular method to
obtain imputations is to use the `Multiple Imputation by Chained
Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an iterative
algorithmic procedure to draw imputations from the posterior predictive
distribution of the missing values. This introduces a potential threat
to the validity of the imputations: what if the algorithm has not
converged? Are the implications then to be trusted? And can we rely on
the inference obtained on the completed data? These are all open
questions, because the convergence properties of iterative imputation
algorithms have not been systematically studied \citep{buur18}.
Moreover, there is no scientific consensus on the convergence properties
of MI algorithms \citep{taka17}. Some default MICE techniques (e.g.,
`predictive mean modeling') might not yield converged states at all
\citep{murr18}. Therefore, algorithmic convergence should be monitored
carefully.

The recommended practice to monitor convergence is to visually inspect
imputations for signs of non-convergence. This practice is insufficient
on two counts: 1) it may be challenging to the untrained eye, and 2)
only severely pathological cases of non-convergence may be diagnosed
\citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic
evaluation of convergence would be preferred.

Iterative imputation algorithms such as MICE are Markov chain Monte
Carlo (MCMC) methods, which means that convergence is not from a scalar
to a point but from one distribution to another. The values generated by
the algorithm (e.g., imputed values) will vary even after convergence.
Therefore, the aim of convergence diagnostics for MCMC methods should
not be to establish the point at which convergence is reached, but to
monitor signs of non-convergence. Several convergence diagnostics for
MCMC methods exist, but it is not known whether these are appropriate
for MICE.

\hypertarget{research-question-how-and-when-to-diagnose-non-convergence}{%
\subsubsection{Research question: how and when to diagnose non
convergence?}\label{research-question-how-and-when-to-diagnose-non-convergence}}

Sub-questions: How can non-convergence be diagnosed? Are common MCMC
non-convergence diagnostics appropriate for MICE? And if so, which
threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{(These are part of the methods, not
RQ per se: What are the effects of continued iterations on estimates,
predictions and inferences? And finally, do the answers differ with
varying missingness proportions? That is, we vary the nr of iterations
and the missingness proportion because we assume that the algorithm has
not reached convergence at \(t=1\), and performs worse with moderate to
high missingness (not so much with little or a LOT of missingness).)}

For reasons of brevity, we only focus on the MI algorithm implemented in
the world-leading MI software: the \texttt{R} \citep{R} package
\texttt{mice} \citep{mice}. The convergence properties of this MI
algorithm are investigated through model-based simulation. The results
of this simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets.

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

@@

\hypertarget{convergence-diagnostics}{%
\section{Convergence diagnostics}\label{convergence-diagnostics}}

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}, see Figure \ref{fig:non-conv}. Without
mixing, imputation chains do not intermingle nicely. Without
stationarity, there is trending within imputation chains. While the
current practice is to inspect traceplots for signs of non-mixing and
non-stationarity, diagnostic evaluation of these two challenges would be
preferred. Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used.

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains.

Positive \(AC\)-values indicate recurrence. If \(\theta\)-values of
subsequent iterations are similar, trending may occur. Negative
\(AC\)-values show no threat to the stationarity component of
convergence. On the contrary even---negative \(AC\)-values indicate that
\(\theta\)-values of subsequent iterations diverge from one-another,
which may increase the variance of \(\theta\) and speed up convergence.
As convergence diagnostic, the interest is therefore in positive
\(AC\)-values. Moreover, the magnitude of \(AC\)-values may be evaluated
statistically, but that is outside of this note's scope.

Negative \(AC\)-values indicate divergence within imputation chains.
\textbf{Subsequent sampled values within each imputation chain are less
alike.}

In short, convergence is reached when there is no dependency between
subsequent iterations of imputation chains (\(AC = 0\)), and chains
intermingle such that the only difference between the chains is caused
by the randomness induced by the algorithm (\(\widehat{R} = 1\)).

@@

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
be too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values
are chosen randomly from the observed data. Therefore, we cannot be
certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\)
\textbf{explain that the randomness induced by the MICE algorithm would
take care of this??}. \textbf{(Add actual hypothesis: we thought that
mice would converge sooner than \(\widehat{R}\) would indicate that it
did)} No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic. \textbf{Added later: High \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.}

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \textbf{someone} (2002), we
will use introduce a \(\theta\) that summarizes the multivariate state
space of the algorithm. Namely, the first eigenvalue of the
variance-covariance matrix of the \(M\) completed datasets.

@@

@@

For empirical researchers: 1) Check trace plots for pathological non
convergence and adjust imputation model if necessary. 2) use
\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
\textbf{?} as threshold . Keep iterating until these thresholds are
reached. 3) Do not use the R function ACF. Instead compute
autocorrelations manually , see for example GitHub link. 4) Track your
own scalar summary of interest. This is advanced but explained in Van
Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this
scalar summary.

@@

Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations,
not iterations (unique for imp data??)

If I don't include it in my study, also: Significance of \(AC\) values;
Estimate (PCA or reg. coeff.) versus \(\widehat{R}\) and \(AC\) on the
estimate

>>>>

\title{Missing the Point: Non-Convergence in Iterative Imputation Algorithms}

@@

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous, but also tedious. If a dataset
contains just one incomplete observation, statistical inferences are
undefined and will not produce any results. To circumvent this, many
statistical packages employ list-wise deletion by default (i.e.,
ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
solution may yield wildly invalid results \citep{buur18}. An alternative
is to `impute' (i.e., fill in) the missing values in the incomplete
observations. Subsequently, statistical inferences can be performed on
the completed dataset. By repeating this process several times, a
distribution of plausible results may be obtained, which reflects the
uncertainty in the data due to missingness. This technique is known as
`multiple imputation' \citep[MI;][]{rubin76}. MI yields unbiased and
confidence valid estimates of the true---but missing---data inference
under many circumstances \citep{buur18}.

\textbf{Move this after terminology??} Figure \ref{fig:diagram} provides
an overview of the steps involved with MI---from incomplete data, to
\(M\) completed datasets, to \(M\) estimated quantities of interest
(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
\(M\) completed datasets. On each completed dataset, the analysis of
scientific interest (the `substantive model', or `complete data model')
is performed. The quantity of scientific interest (e.g., a regression
coefficient) is denoted with \(Q\). Since \(Q\) is estimated on each
completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained. These
\(M\) values are combined into a single pooled estimate \(\bar{Q}\).
\textbf{ALSO: make clear that iteration happens within the first step,
between incomplete and completed data.}

@@

A popular method to obtain imputations is to use the `Multiple
Imputation by Chained Equations' algorithm, shorthand
`MICE'\citep{mice}. MICE is an iterative algorithmic procedure to draw
imputations from the posterior predictive distribution of the missing
values. This introduces a potential threat to the validity of the
imputations: What if the algorithm has not converged? Are the
implications then to be trusted? And can we rely on the inference
obtained on the completed data? These are all open questions, because
the convergence properties of iterative imputation algorithms have not
been systematically studied \citep{buur18}. Moreover, there is no
scientific consensus on the convergence properties of MI algorithms
\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
modeling') might not yield converged states at all \citep{murr18}.
Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating convergence is to
visually inspect imputations for signs of non-convergence. This method
is insufficient on two counts: 1) it may be challenging to the untrained
eye, and 2) only severely pathological cases of non-convergence may be
diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
diagnostic evaluation of convergence would be preferred.

Monitoring convergence of iterative imputation algorithms diagnostically
is challenging. Iterative imputation algorithms such as MICE are Markov
chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
from a scalar to a point but from one distribution to another. The
values generated by the algorithm (e.g., imputed values) will vary even
after convergence. Therefore, the aim of convergence diagnostics for
MCMC methods is not to establish the point at which convergence is
reached, but to monitor signs of non-convergence. Several of such
diagnostics exist for MCMC methods, but it is not known whether these
are appropriate for MICE.

In this paper we investigate how non-convergence in iterative imputation
algorithms may be diagnosed, how well these methods perform, and at
which point convergence may safely be assumed. For reasons of brevity,
we only focus on the MI algorithm implemented in the world-leading MI
software: the \texttt{R} \citep{R} package \texttt{mice} \citep{mice}.
The convergence properties of this MI algorithm are investigated through
model-based simulation. The results of this simulation study are
guidelines for assessing convergence of MI algorithms, which will aid
applied researchers in drawing valid inference from incomplete datasets.

\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
Are common MCMC non-convergence diagnostics appropriate for MICE? And if
so, which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)?

\hypertarget{terminology}{%
\subsection{Terminology}\label{terminology}}

@@

\hypertarget{how-may-non-convergence-be-diagnosed}{%
\section{How may non-convergence be
diagnosed?}\label{how-may-non-convergence-be-diagnosed}}

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}. Without mixing, imputation chains do
not intermingle nicely. Without stationarity, there is trending within
imputation chains, see Figure \ref{fig:non-conv}. \textbf{Explain what
we see, namely example by \citet{buur18} reproduced, showing the
traceplots of chain means for some variable. The first plot is typical
convergence of MICE, the second is pathological non-convergence because
of a mis-specified imputation model. Each line is an imputation. In the
first plot, the lines intermingle nicely and there is little trending.
In the second, there is a lot of trending and the lines do not cross
much. Importantly, the chain means at the last iteration (the imputed
value per \(m\)) are very different between the two plots. The algorithm
with the mis-specified model yields imputed values that are on average a
magnitude two larger than those of the typically converged algorithm.
This shows the importance of reaching convergend states in iterative
imputation algorithms.}

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used.

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one-another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
evaluated statistically, but that is outside of this note's scope.

\hypertarget{in-practiceanswer}{%
\subsection{In practice/Answer}\label{in-practiceanswer}}

\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence diagnostics for MICE algorithms, the following condtion
must hold when applied on the two algorithms plotted above: the methods
indicate worse performance for the mis-specified model with pathological
non-convergence (i.e., higher \(\widehat{R}\)- and \(AC\)-values than
the typical performance). And additionally the methods should reflect
the increasing convergence in the typical convergence situation as the
number of iterations goes up.}

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together. So are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up in the typical convergence situation. The superior
performance of the typical convergence over the pathological
non-convergnce is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function stats::acf(). When
we look at this panel, we must conclude that something weird is going
on. The \(AC\)-values indicate equal performance (upto \(t=5\)) for the
typical convergence and the pathological non-convergnce, while there is
obvious trending in the latter. Moreover, the best convergence (as
indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending upto iteration number seven. After consulting the documentation
on stats::acf(), we conclude that this implementation of \(AC\) is not
suitable for iterative imputation algorithms. There is a correction
factor for a mathematical shortcut that works in the limit. According to
\ldots{} the function works when the number of iterations \textgreater{}
50. The default number of iterations in mice, however, is 5. Therefore,
we compute \(AC\) manually, see panel D. The AC values in this plot do
meet the requirements (\ldots) and will therefore be used in the
simulation study.}

@@

Convergence is reached when there is no dependency between subsequent
iterations of imputation chains (\(AC = 0\)), and chains intermingle
such that the only difference between the chains is caused by the
randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
that a completely convergenced state will not be reached.

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
be too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values
are chosen randomly from the observed data. Therefore, we cannot be
certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\)
\textbf{explain that the randomness induced by the MICE algorithm would
take care of this??}. \textbf{(Add actual hypothesis: we thought that
mice would converge sooner than \(\widehat{R}\) would indicate that it
did)} No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic. \textbf{Added later: High \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.
But this doesn't seem the case in practice??}

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \citep{mack03}, we will use
introduce a \(\theta\) that summarizes the multivariate state space of
the algorithm. Namely, the first eigenvalue of the variance-covariance
matrix of the \(M\) completed datasets.

@@

{\centering \includegraphics{manuscript_files/figure-latex/mean-1}

\caption{Convergence diagnostics chain mean.}\label{fig:mean}
\end{figure}

{\centering \includegraphics{manuscript_files/figure-latex/sd-1}

\caption{Convergence diagnostics chain variance.}\label{fig:sd}
\end{figure}
\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/est-1}

\caption{Convergence diagnostics regression coefficient.}\label{fig:est}
\end{figure}

{\centering \includegraphics{manuscript_files/figure-latex/pred-1}

\caption{Convergence diagnostics regression coefficient.}\label{fig:pred}
\end{figure}

@@

\textbf{For empirical researchers: 1) Check trace plots for pathological
non convergence and adjust imputation model if necessary. 2) use
\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with }?** as
threshold . Keep iterating until these thresholds are reached. 3) Do not
use the R function ACF. Instead compute autocorrelations manually , see
for example GitHub link. 4) Track your own scalar summary of interest.
This is advanced but explained in Van Buren 2018. Compute
\(\widehat{R}\) and autocorrelation values for this scalar summary. **

@@

\textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
imputations, not iterations (unique for imp data??)}

\textbf{If I don't include it in my study, also: Significance of \(AC\)
values}
